---
title: "Estimating Transcription Rates with STADyUM from PRO-seq Data"
author: "Xin Zeng"
date: "2026/04/01"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: default
    highlight: tango
---

# Introduction

This tutorial demonstrates how to use the **STADyUM** package to estimate transcription rates from PRO-seq data. STADyUM (Statistical Analysis of Dynamics Under Multiple Measurements) models RNA polymerase dynamics by analyzing reads at promoter-proximal pause sites and gene bodies. 

In this workflow, we will:
1. Harmonize transcription start sites (TSSs) between two cell types (CD4 and CD14)
2. Construct read-counting regions for pause sites and gene bodies
3. Apply STADyUM to estimate key kinetic parameters (release rate, elongation rate, activity)
4. Visualize the relationships between estimated transcription parameters

This example uses human PRO-seq data from two immunologically distinct cell types to demonstrate comparative transcription rate estimation.

---

# Setup

## Load Required Libraries

```{r setup, message=FALSE, warning=FALSE}
library(STADyUM)           # For transcription rate estimation
library(GenomicRanges)     # For genomic data manipulation
library(dplyr) 
library(plyranges)        
#library(tidyverse)         # For data manipulation and visualization
library(ggplot2)           # For plotting
library(pracma)
library(BiocFileCache)
```

```{r file-paths}

## PULL ALL DATA FROM GEO ##

# Required by Bioconductor to check for connection
if (!requireNamespace("BiocFileCache", quietly = TRUE) || 
    !curl::has_internet()) {
  knitr::opts_chunk$set(eval = FALSE)
  message("No internet connection; vignette code not evaluated.")
}

bfc <- BiocFileCache(ask = FALSE)

#geo_base <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSEnnn/GSEnnnn/suppl"
#tsn_path <- bfcrpath(bfc, file.path(geo_base, "PROseq-HUMAN-CD4_tsn.RDS"))

# TSN (Transcription Start Site Network) files
# These are GRanges objects containing TSS annotations for each sample
cd4_tsn_file <- paste0(root_dir, 'inst/extdata/lrts/PROseq-HUMAN-CD4_tsn.RDS')
cd14_tsn_file <- paste0(root_dir, 'inst/extdata/lrts/PROseq-HUMAN-CD14_tsn.RDS')

# Transcript annotation file
# Contains gene models with exon structure for all genes
human_tq_file <- paste0(root_dir, 'inst/extdata/lrts/HUMAN.RDS')

# PRO-seq signal files (stranded bigWig format)
# Plus strand reads
cd4_bw_plus <- paste0(root_dir, 'data/result2/bigwig/p3/PROseq-HUMAN-CD4_plus.bw')
cd4_bw_minus <- paste0(root_dir, 'data/result2/bigwig/p3/PROseq-HUMAN-CD4_minus.bw')

cd14_bw_plus <- paste0(root_dir, 'data/result2/bigwig/p3/PROseq-HUMAN-CD14_plus.bw')
cd14_bw_minus <- paste0(root_dir, 'data/result2/bigwig/p3/PROseq-HUMAN-CD14_minus.bw')

```

---

# Data Loading and TSS Selection

## Load TSS Annotations

We begin by loading transcription start site (TSS) annotations from both cell types. These annotations may contain multiple TSSs per gene (alternative promoters). For this analysis, we select a single representative TSS per gene.

```{r load-tss}
# Load TSN data for both cell types
cd4_tsn <- readRDS(cd4_tsn_file)
cd14_tsn <- readRDS(cd14_tsn_file)

# Display structure of TSN data
head(cd4_tsn)
```

## Select Single TSS per Gene

When multiple TSSs exist for a single gene, we retain only the most upstream TSS. This ensures consistent TSS usage across the analysis. The "most upstream" is defined as the TSS with the smallest genomic coordinate on the plus strand and the largest coordinate on the minus strand.

```{r select-single-tss}
# Process CD4 data: select most upstream TSS per gene
cd4_tsn <- as.data.frame(cd4_tsn) %>%
  group_by(ensembl_gene_id, strand) %>%
  slice_min(order_by = ifelse(strand == "+", start, -start), 
            n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  as_granges()

# Process CD14 data: select most upstream TSS per gene
cd14_tsn <- as.data.frame(cd14_tsn) %>%
  group_by(ensembl_gene_id, strand) %>%
  slice_min(order_by = ifelse(strand == "+", start, -start), 
            n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  as_granges()

cat("CD4 TSS sites:", length(cd4_tsn), "\n")
cat("CD14 TSS sites:", length(cd14_tsn), "\n")
```

---

# TSS Harmonization Between Samples

## Merge TSSs from Both Cell Types

When comparing transcription rates between cell types, TSSs should be unified to ensure we measure the same transcriptional unit. If TSSs are close between samples (within 30 bp), we merge them and use the most upstream as the consensus TSS.

```{r harmonize-tss}
# Merge TSSs: combine close TSSs from both samples
# TSSs within 30 bp are unified to the most upstream position
tsn_merge <- STADyUM:::merge_tss_from_two_samples(cd4_tsn, cd14_tsn)

# Extract distances between matched TSSs
dist_df <- data.frame(Distance = tsn_merge$dist)

# Display summary statistics
summary(dist_df$Distance)
```

## Visualize TSS Distance Distribution

This plot shows the distribution of distances between TSSs in CD4 and CD14 samples. The dashed vertical line indicates the 30 bp threshold used for merging.

```{r plot-tss-distances, fig.width=4, fig.height=3}
p0 <- ggplot(dist_df, aes(x = log10(Distance + 1))) +
  geom_histogram(bins = 100, fill = "steelblue", color = "black") +
  scale_y_log10() +
  xlab('log10(Distance + 1)') +
  ylab('Count') +
  geom_vline(xintercept = log10(30 + 1),
             linetype = "dashed",
             color = "red", linewidth = 1)

print(p0)
```

---

# Building Read-Count Regions

## Load Transcript Annotations

We load the full transcript annotation to define gene bodies. From these, we extract the transcription termination sites (TTS) to define the 3' boundary of gene bodies.

```{r load-transcripts}
# Load transcript model for all genes
human_tq <- readRDS(human_tq_file)

# Extract gene ranges and reduce overlapping exons
# This creates a single range per gene (union of all exons)
gngrng <- human_tq@transcripts %>%
  group_by(ensembl_gene_id) %>%
  reduce_ranges_directed() %>%
  sort()

cat("Number of genes:", length(gngrng), "\n")
```

## Define Pause and Gene Body Regions

For STADyUM analysis, we define two counting regions:
- **Pause region**: typically near the TSS (promoter-proximal pause site)
- **Gene body region**: the elongating polymerase region 

Here, we anchor at the transcription termination site (3' end) and define regions relative to those anchors.

```{r define-count-regions}
# Anchor at 3' end (transcription termination site) and set width to 1
# This creates a point coordinate at the TTS for each gene
bw_tts <- gngrng %>% 
  plyranges::anchor_3p() %>% 
  mutate(width = 1)

# Build read-count regions for CD4
# Returns list with 'pause' and 'gene_body' GRanges
cd4_count_region <- STADyUM:::build_readcount_regions(tsn_merge$tsn1, bw_tts)

# Build read-count regions for CD14
cd14_count_region <- STADyUM:::build_readcount_regions(tsn_merge$tsn2, bw_tts)

cat("CD4 pause regions:", length(cd4_count_region$pause), "\n")
cat("CD4 gene body regions:", length(cd4_count_region$gene_body), "\n")
```

---

# Running STADyUM

## Estimate Transcription Rates

The `estimateTranscriptionRates()` function applies STADyUM to count reads in pause and gene body regions. It estimates key kinetic parameters:

- **beta (β)**: pause release rate (1/time)
- **chi (χ)**: elongation rate relative to transcription activity
- **activity**: transcription initiation/activity level

```{r run-stadyum}
# Estimate transcription rates for CD4
# Input: plus and minus strand bigWig files, pause/gene body regions, sample label
cd4_rate <- estimateTranscriptionRates(
  cd4_bw_plus, cd4_bw_minus,
  cd4_count_region$pause, cd4_count_region$gene_body, 
  "CD4"
)

# Estimate transcription rates for CD14
cd14_rate <- estimateTranscriptionRates(
  cd14_bw_plus, cd14_bw_minus,
  cd14_count_region$pause, cd14_count_region$gene_body, 
  "CD14"
)

cat("CD4 rate estimation complete\n")
cat("CD14 rate estimation complete\n")
```

---

# Post-Processing Results

## Process Estimated Rates

After STADyUM estimation, rates are post-processed to remove outliers and standardize values. This typically involves log-transformation and filtering of extreme values.

```{r post-process-rates}
# Post-process rates for CD4
# This filters outliers and applies statistical normalization
cd4_rate@rates <- STADyUM:::process_sample_rate(rates(cd4_rate))

# Post-process rates for CD14
cd14_rate@rates <- STADyUM:::process_sample_rate(rates(cd14_rate))

cat("Post-processing complete\n")
```

## Save Results

Save the estimated rates as RDS files for downstream analysis.

```{r save-rates}
# Save CD4 rates
saveRDS(cd4_rate, paste0(file.path(result_dir, "human/cd4_rate"), ".RDS"))

# Save CD14 rates
saveRDS(cd14_rate, paste0(file.path(result_dir, "human/cd14_rate"), ".RDS"))

cat("Results saved to:", result_dir, "\n")
```

---

# Visualization

## Extract Rate Data

Convert rate objects into data frames for plotting.

```{r extract-rates}
# Create a CD14 visualization copy with extreme outliers removed
cd14_rate_viz <- cd14_rate
cd14_rate_viz@rates <- cd14_rate_viz@rates[log10(cd14_rate_viz@rates$betaAdp) < 10, ]

cat("CD4 genes with estimated rates:", nrow(rates(cd4_rate)), "\n")
cat("CD14 genes with estimated rates:", nrow(rates(cd14_rate_viz)), "\n")
```

## CD4 Visualizations

### Pause Release Rate vs. Transcription Activity

This plot shows the relationship between the pause release rate (beta) and transcription activity (chi). Each point represents one gene. Positively correlated genes may indicate coupled regulation of pause release and transcription initiation.

```{r cd4-beta-chi, fig.width=3.5, fig.height=3}
p1 <- plotScatterDensity(cd4_rate, "betaAdp", "chi",
                          log_x = TRUE, log_y = TRUE,
                          xlab = expression(log[10] ~ beta),
                          ylab = expression(log[10] ~ chi),
                          xlim = c(-6, -1), ylim = c(-4, 1))

if (requireNamespace("ggpubr", quietly = TRUE)) {
  p1 <- p1 + scale_color_viridis_c() +
    ggpubr::stat_cor(method = "spearman", label.x = -3.5, label.y = -3.8,
                     size = 5, aes(label = after_stat(r.label)))
}

print(p1)
```

### Pause Release Rate vs. Elongation Parameters

These plots show relationships between pause release rate (beta) and two measures of elongation variability. The first shows beta vs. mean elongation time; the second shows beta vs. elongation time standard deviation.

```{r cd4-beta-elongation, fig.width=10, fig.height=3}
p1 <- plotScatterDensity(cd4_rate, "betaAdp", "fkMean",
                          log_x = TRUE,
                          xlab = expression(log[10] ~ beta),
                          ylab = expression(mu))
if (requireNamespace("ggpubr", quietly = TRUE)) {
  p1 <- p1 + scale_color_viridis_c() +
    ggpubr::stat_cor(method = "spearman", label.x = -2.2, label.y = 160,
                     size = 5, aes(label = after_stat(r.label)))
}

p2 <- plotScatterDensity(cd4_rate, "betaAdp", "fkSD",
                          log_x = TRUE,
                          xlab = expression(log[10] ~ beta),
                          ylab = expression(sigma),
                          color_var = "sdGroup",
                          color_values = c(Broad = "#1b9e77", Sharp = "#762a83"),
                          color_lab = expression(sigma ~ group))
if (requireNamespace("ggpubr", quietly = TRUE)) {
  p2 <- p2 +
    ggpubr::stat_cor(method = "spearman", label.x = -2.5, label.y = 75,
                     size = 3, aes(label = after_stat(r.label), group = 1),
                     color = "black")
}
p2_hist <- ggplot(rates(cd4_rate), aes(x = .data$fkSD)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50) +
  geom_vline(xintercept = 38.5, linetype = "dashed", color = "black",
             linewidth = 0.6) +
  coord_flip() +
  theme(axis.title = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank())
print(p1)

grid::grid.newpage()
grid::pushViewport(grid::viewport(layout = grid::grid.layout(1, 2, widths = grid::unit(c(4, 1), "null"))))
print(p2,      vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 1))
print(p2_hist, vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 2))

p3 <- plotScatterDensity(cd4_rate, "fkMean", "fkSD",
                          xlab = expression(mu),
                          ylab = expression(sigma))
if (requireNamespace("ggpubr", quietly = TRUE)) {
  p3 <- p3 + scale_color_viridis_c() +
    ggpubr::stat_cor(method = "spearman", label.x = 112, label.y = 1,
                     size = 5, aes(label = after_stat(r.label)))
}

print(p3)

```

## CD14 Visualizations

### Pause Release Rate vs. Transcription Activity

Same analysis for the CD14 cell type. This allows comparison of transcription parameter distributions and relationships between the two cell types.

```{r cd14-beta-chi, fig.width=3.5, fig.height=3}
p1 <- plotScatterDensity(cd14_rate_viz, "betaAdp", "chi",
                          log_x = TRUE, log_y = TRUE,
                          xlab = expression(log[10] ~ beta),
                          ylab = expression(log[10] ~ chi),
                          xlim = c(-6, -1), ylim = c(-4, 1))

if (requireNamespace("ggpubr", quietly = TRUE)) {
  p1 <- p1 + scale_color_viridis_c() +
    ggpubr::stat_cor(method = "spearman", label.x = -3, label.y = -4,
                     size = 5, aes(label = after_stat(r.label)))
}

print(p1)
```

### Pause Release Rate vs. Elongation Parameters

Relationship plots for CD14 showing how pause release rate associates with elongation time parameters.

```{r cd14-beta-elongation, fig.width=10, fig.height=3}
p1 <- plotScatterDensity(cd14_rate_viz, "betaAdp", "fkMean",
                          log_x = TRUE,
                          xlab = expression(log[10] ~ beta),
                          ylab = expression(mu))
if (requireNamespace("ggpubr", quietly = TRUE)) {
  p1 <- p1 + scale_color_viridis_c() +
    ggpubr::stat_cor(method = "spearman", label.x = 0, label.y = 180,
                     size = 5, aes(label = after_stat(r.label)))
}

p2 <- plotScatterDensity(cd14_rate_viz, "betaAdp", "fkSD",
                          log_x = TRUE,
                          xlab = expression(log[10] ~ beta),
                          ylab = expression(sigma),
                          color_var = "sdGroup",
                          color_values = c(Broad = "#1b9e77", Sharp = "#762a83"),
                          color_lab = expression(sigma ~ group))
if (requireNamespace("ggpubr", quietly = TRUE)) {
  p2 <- p2 +
    ggpubr::stat_cor(method = "spearman", label.x = 0, label.y = 75,
                     size = 3, aes(label = after_stat(r.label), group = 1),
                     color = "black")
}
p2_hist <- ggplot(rates(cd14_rate_viz), aes(x = .data$fkSD)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50) +
  geom_vline(xintercept = 37.5, linetype = "dashed", color = "black",
             linewidth = 0.6) +
  coord_flip() +
  theme(axis.title = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank())
print(p1)

grid::grid.newpage()
grid::pushViewport(grid::viewport(layout = grid::grid.layout(1, 2, widths = grid::unit(c(4, 1), "null"))))
print(p2,      vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 1))
print(p2_hist, vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 2))

p3 <- plotScatterDensity(cd14_rate_viz, "fkMean", "fkSD",
                          xlab = expression(mu),
                          ylab = expression(sigma))
if (requireNamespace("ggpubr", quietly = TRUE)) {
  p3 <- p3 + scale_color_viridis_c() +
    ggpubr::stat_cor(method = "spearman", label.x = 120, label.y = 1,
                     size = 5, aes(label = after_stat(r.label)))
}

print(p3)

```

---

# Summary

This tutorial demonstrated a complete workflow for estimating transcription rates from PRO-seq data using STADyUM:

1. **TSS Harmonization**: We unified transcription start sites between two cell types, allowing consistent measurement of transcriptional parameters across samples.

2. **Region Definition**: We defined pause (promoter-proximal) and gene body regions as input for rate estimation.

3. **Rate Estimation**: STADyUM estimated three key transcription parameters (pause release rate, elongation characteristics, and activity) from the read distribution in these regions.

4. **Visualization**: We explored relationships between parameters, revealing how pause release and elongation dynamics are coupled in CD4 and CD14 cells.

The estimated rates are now ready for downstream comparative analysis (e.g., identifying genes with cell-type-specific transcription dynamics) and integration with other genomic data.

---

# Session Information

```{r session-info}
sessionInfo()
```
