---
title: "Comparative Analysis of Promoter-Proximal Pausing using Likelihood Ratio Tests"
author: "Xin Zeng"
date: "2026/04/01"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: default
    highlight: tango
vignette: >
  %\VignetteIndexEntry{Comparative Analysis of Promoter-Proximal Pausing using Likelihood Ratio Tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

This tutorial demonstrates how to use the **STADyUM** package to perform comparative analysis of pause-escape kinetics and pause-site distributions across cell types using the likelihood-ratio tests (LRTs) developed in [CITE]. 

In this workflow, we will:
1. 

This example uses human PRO-seq data from two immunologically distinct cell types to demonstrate comparative transcription rate estimation.

---

# Setup

## Load Required Libraries and Data

```{r setup, message=FALSE, warning=FALSE}
library(STADyUM)           # For transcription rate estimation
library(GenomicRanges)     # For genomic data manipulation
library(dplyr) 
library(plyranges)        
#library(tidyverse)         # For data manipulation and visualization
library(ggplot2)           # For plotting
library(pracma)
#library(BiocFileCache)
```

```{r file-paths}

## PULL ALL DATA FROM ZENODO ##

# Required by Bioconductor to check for connection
if (!requireNamespace("BiocFileCache", quietly = TRUE) || 
    !curl::has_internet()) {
  knitr::opts_chunk$set(eval = FALSE)
  message("No internet connection; vignette code not evaluated.")
}

#bfc <- BiocFileCache(ask = FALSE)

#geo_base <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSEnnn/GSEnnnn/suppl"
#tsn_path <- bfcrpath(bfc, file.path(geo_base, "PROseq-HUMAN-CD4_tsn.RDS"))

# STADyUM `ExperimentTranscriptionRates` objects containing transcription rates estimated from human CD4+ T cell and human CD14+ monocyte PRO-seq data.
cd4_rate <- 'lrt_test_data/cd4_rate.RDS'
cd14_rate <- 'lrt_test_data/cd14_rate.RDS'
scale_factor <- 38359083 / 33294862 

```

---

# Running STADyUM

## Likelihood Ratio Test Calculations

The `estimateTranscriptionRates()` function applies STADyUM to count reads in pause and gene body regions. It estimates key kinetic parameters:

- **beta (β)**: pause release rate (1/time)
- **chi (χ)**: elongation rate relative to transcription activity
- **activity**: transcription initiation/activity level

```{r lrt_calculation}

lrt <- likelihoodRatioTest(cd4_rate, cd14_rate, scale_factor)

lrt <- mergeBetaChiLRTStats(lrt, scale_factor, prefix1='CD4', prefix2='CD14')
lrt$pause_change <- paste0(lrt$CD4_sdGroup, '_to_', lrt$CD14_sdGroup)

lrt$deltaSD   <- (lrt$CD14_fkSD - lrt$CD4_fkSD) /
                      (lrt$CD14_fkSD + lrt$CD4_fkSD)
lrt$deltaMean <- lrt$fkMean2 - lrt$fkMean1

head(lrt)

cat("CD4 rate estimation complete\n")
cat("CD14 rate estimation complete\n")
```

---

# Visualization
```{r lrt_visualization}


```

---

# Summary

This tutorial demonstrated a complete workflow for estimating transcription rates from PRO-seq data using STADyUM:

1. **TSS Harmonization**: We unified transcription start sites between two cell types, allowing consistent measurement of transcriptional parameters across samples.

2. **Region Definition**: We defined pause (promoter-proximal) and gene body regions as input for rate estimation.

3. **Rate Estimation**: STADyUM estimated three key transcription parameters (pause release rate, elongation characteristics, and activity) from the read distribution in these regions.

4. **Visualization**: We explored relationships between parameters, revealing how pause release and elongation dynamics are coupled in CD4 and CD14 cells.

The estimated rates are now ready for downstream comparative analysis (e.g., identifying genes with cell-type-specific transcription dynamics) and integration with other genomic data.

---

# Session Information

```{r session-info}
sessionInfo()
```
