---
title: "STADyUM: Simulating and Analyzing Transcription Dynamics"
author: "Rebecca Hassett"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{STADyUM: Simulating and Analyzing Transcription Dynamics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r load_package, include = FALSE}
# Load the package
library(STADyUM)

```

# Introduction

STADyUM is an R package for simulating and analyzing transcription dynamics. 
It provides tools for:
- Simulating polymerase movement along genes
- Estimating transcription rates
- Analyzing pause site distributions
- Visualizing transcription dynamics

# Installation

```{r install, eval = FALSE}
# Install from GitHub
devtools::install_github("yourusername/STADyUM")
```

# Basic Usage

## Estimating Transcription Rates from Experimental Data

## Creating a Simulate Polymerase Object

First, let's create a simulate_polymerase object with some parameters:

```{r create_simpol}
library(STADyUM)
```

# Session Info

```{r session_info}
sessionInfo()
``` 
