---
title: "STADyUM: Simulating and Analyzing Transcription Dynamics"
author: "Rebecca Hassett"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{STADyUM: Simulating and Analyzing Transcription Dynamics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r load_package, include = FALSE}
# Load the package
library(STADyUM)
```

# Introduction

STADyUM is an R package for simulating and analyzing transcription dynamics. 
It provides tools for:
- Simulating polymerase movement along genes
- Estimating transcription rates
- Analyzing pause site distributions
- Visualizing transcription dynamics

# Installation

```{r install, eval = FALSE}
# Install from GitHub
devtools::install_github("yourusername/STADyUM")
```

# Basic Usage

## Estimating Transcription Rates from Experimental Data

## Creating a Simulate Polymerase Object

First, let's create a simulate_polymerase object with some parameters:

```{r create_simpol}
library(STADyUM)

# Create a simulate_polymerase object
simpol <- simulatePolymerase(
    k = 50, # mean pause sites
    ksd = 25, # standard deviation of pause sites
    kMin = 17, # minimum pause sites
    kMax = 200, # maximum pause sites
    geneLen = 1950, # gene length
    alpha = 1, # initiation rate
    beta = 1, # pause release rate
    zeta = 2000, # mean elongation rate
    zetaSd = 1000, # standard deviation of elongation rate
    zetaMin = 1500, # minimum elongation rate
    zetaMax = 2500, # maximum elongation rate
    cellNum = 1000, # number of cells
    polSize = 33, # polymerase size
    addSpace = 17, # additional space
    time = 1, # simulation time
    stepsToRecord = 1 # steps to record
)
```

## Estimating Transcription Rates

Now we can estimate transcription rates from the simulation:

```{r estimate_rates}
# Estimate transcription rates
srates <- estimateSimulationTranscriptionRates(simpol)

# Get the rates
chi <- chi(srates)
beta_org <- betaOrg(srates)
beta_adp <- betaAdp(srates)
phi <- phi(srates)
```

## Visualizing Results

The package provides several plotting functions:

```{r plotting}
# Plot transcription rates
plotTranscriptionRates(srates)
```

# Advanced Usage

## Steric Hindrance

You can enable steric hindrance calculations:

```{r steric_hindrance}
# Estimate rates with steric hindrance
srates_sh <- estimateSimulationTranscriptionRates(simpol,
    stericHindrance = TRUE
)

# Get steric hindrance values
phi_sh <- phi(srates_sh)
```

## Saving and Loading Results

You can save your results to RDS files:

```{r save_load}
# Save results
saveRDS(srates, "simulation_results.RDS")

# Load results
srates_loaded <- readRDS("simulation_results.RDS")
```

## Accessing Simulation Parameters

You can access the original simulation parameters:

```{r parameters}
# Get parameters
params <- parameters(simpol)

# Print parameters
print(params)
```

# Session Info

```{r session_info}
sessionInfo()
``` 
