---
title: "STADyUM: Simulating and Analyzing Transcription Dynamics"
author: "Rebecca Hassett"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{STADyUM: Simulating and Analyzing Transcription Dynamics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Introduction

STADyUM is an R package for simulating and analyzing transcription dynamics. It provides tools for:
- Simulating polymerase movement along genes
- Estimating transcription rates
- Analyzing pause site distributions
- Visualizing transcription dynamics

# Installation

```{r install, eval = FALSE}
# Install from GitHub
devtools::install_github("yourusername/STADyUM")
```

# Basic Usage

## Estimating Transcription Rates from Experimental Data

```{r estimate_rates_from_experimental_data}
load("experimental_data/granges_for_read_counting.RData")

exp_rates_control <- estimate_experiment_transcription_rates(
    bigwig_plus = "experimental_data/PROseq-K562-vihervaara-control-SE_plus.bw",
    bigwig_minus = "experimental_data/PROseq-K562-vihervaara-control-SE_minus.bw",
    pause_regions = bw_pause_filtered,
    gene_body_regions = bw_gb_filtered,
    gene_name_column = "gene_id",
    steric_hindrance = TRUE,
    omega_scale = 1.62776148862949
)

exp_rates_treatment <- estimate_experiment_transcription_rates(
    bigwig_plus = "experimental_data/PROseq-K562-vihervaara-treatment-SE_plus.bw",
    bigwig_minus = "experimental_data/PROseq-K562-vihervaara-treatment-SE_minus.bw",
    pause_regions = bw_pause_filtered,
    gene_body_regions = bw_gb_filtered,
    gene_name_column = "gene_id",
    steric_hindrance = TRUE,
    omega_scale = 1.62776148862949
)

# Create a data frame for comparison
comparison_df <- data.frame(
    gene_id = c(rates(exp_rates_control)$gene_id, rates(exp_rates_treatment)$gene_id),
    beta_adp = c(rates(exp_rates_control)$beta_adp, rates(exp_rates_treatment)$beta_adp),
    condition = rep(c("Control", "Treatment"), each = nrow(rates(exp_rates_control)))
)

# Create a boxplot
library(ggplot2)
ggplot(comparison_df, aes(x = condition, y = beta_adp, fill = condition)) +
    geom_boxplot() +
    scale_fill_brewer(palette = "Set2") +
    labs(
        title = "Comparison of Pause Release Rates",
        x = "Condition",
        y = "Beta (pause release rate)"
    ) +
    theme_minimal()

# Create a scatter plot of original vs adapted rates
ggplot(rates(exp_rates_control), aes(x = beta_org, y = beta_adp)) +
    geom_point(alpha = 0.6, color = "#1E88E5") +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
    labs(
        title = "Original vs Adapted Pause Release Rates",
        x = "Original Beta",
        y = "Adapted Beta"
    ) +
    theme_minimal()

# Create a density plot of pause site positions
ggplot(rates(exp_rates_control), aes(x = fk_mean)) +
    geom_density(fill = "#1E88E5", alpha = 0.6) +
    labs(
        title = "Distribution of Pause Site Positions",
        x = "Mean Pause Site Position",
        y = "Density"
    ) +
    theme_minimal()

# Create a scatter plot of phi vs beta
ggplot(rates(exp_rates_control), aes(x = beta_adp, y = phi)) +
    geom_point(alpha = 0.6, color = "#1E88E5") +
    labs(
        title = "Relationship between Pause Release Rate and Landing Pad Occupancy",
        x = "Beta (pause release rate)",
        y = "Phi (landing pad occupancy)"
    ) +
    theme_minimal()

# Save the plots
ggsave("comparison_plot.pdf", width = 8, height = 6)
ggsave("scatter_plot.pdf", width = 8, height = 6)
ggsave("pause_sites_plot.pdf", width = 8, height = 6)
ggsave("steric_hindrance_plot.pdf", width = 8, height = 6)
```

## Creating a Simulate Polymerase Object

First, let's create a simulate_polymerase object with some parameters:

```{r create_simpol}
library(STADyUM)

# Create a simulate_polymerase object
simpol <- simulate_polymerase(
    k = 50, # mean pause sites
    k_min = 17, # minimum pause sites
    k_max = 200, # maximum pause sites
    ksd = 25, # standard deviation of pause sites
    gene_len = 1950, # gene length
    alpha = 1, # initiation rate
    beta = 1, # pause release rate
    zeta = 2000, # mean elongation rate
    zeta_sd = 1000, # standard deviation of elongation rate
    zeta_min = 1500, # minimum elongation rate
    zeta_max = 2500, # maximum elongation rate
    total_cells = 20000, # number of cells
    pol_size = 33, # polymerase size
    add_space = 17, # additional space
    time = 40, # simulation time
    steps_to_record = 1 # steps to record
)

simpol2 <- readRDS("simulate_polymerase_object.RDS")
```

## Estimating Transcription Rates

Now we can estimate transcription rates from the simulation:

```{r estimate_rates}
# Estimate transcription rates
srates <- estimate_simulation_transcription_rates(simpol)

# Get the rates
chi <- get_chi(srates)
beta_org <- get_beta_org(srates)
beta_adp <- get_beta_adp(srates)
phi <- get_phi(srates)
```

## Visualizing Results

The package provides several plotting functions:

```{r plotting}
# Plot transcription rates
plot_transcription_rates(srates)

# Plot pause site distribution
plot_pause_site_distribution(srates)

# Plot RNAP counts (if available)
plot_rnap_counts(srates)
```

# Advanced Usage

## Steric Hindrance

You can enable steric hindrance calculations:

```{r steric_hindrance}
# Estimate rates with steric hindrance
srates_sh <- estimate_simulation_transcription_rates(simpol, 
steric_hindrance = TRUE)

# Get steric hindrance values
phi_sh <- get_phi(srates_sh)
```

## Saving and Loading Results

You can save your results to RDS files:

```{r save_load}
# Save results
saveRDS(srates, "simulation_results.RDS")

# Load results
srates_loaded <- readRDS("simulation_results.RDS")
```

## Accessing Simulation Parameters

You can access the original simulation parameters:

```{r parameters}
# Get parameters
params <- get_parameters(simpol)

# Print parameters
print(params)
```

# Session Info

```{r session_info}
sessionInfo()
``` 
