---
title: "STADyUM: Simulating and Analyzing Transcription Dynamics"
author: "Rebecca Hassett"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{STADyUM: Simulating and Analyzing Transcription Dynamics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

STADyUM is an R package for simulating and analyzing transcription dynamics. It provides tools for:
- Simulating polymerase movement along genes
- Estimating transcription rates
- Analyzing pause site distributions
- Visualizing transcription dynamics

# Installation

```{r install, eval = FALSE}
# Install from GitHub
devtools::install_github("yourusername/STADyUM")
```

# Basic Usage

## Creating a Simulate Polymerase Object

First, let's create a simulate_polymerase object with some parameters:

```{r create_simpol}
library(STADyUM)

# Create a simulate_polymerase object
simpol <- simulate_polymerase(
  k = 50,           # mean pause sites
  k_min = 17,       # minimum pause sites
  k_max = 200,      # maximum pause sites
  ksd = 25,         # standard deviation of pause sites
  gene_len = 1950,  # gene length
  alpha = 1,        # initiation rate
  beta = 1,         # pause release rate
  zeta = 2000,      # mean elongation rate
  zeta_sd = 1000,   # standard deviation of elongation rate
  zeta_min = 1500,  # minimum elongation rate
  zeta_max = 2500,  # maximum elongation rate
  total_cells = 20000, # number of cells
  s = 33,           # polymerase size
  h = 17,           # additional space
  time = 40,        # simulation time
  delta_t = 0.0001, # time step
  csv_steps_to_record = 0  # steps to record
)
```

## Estimating Transcription Rates

Now we can estimate transcription rates from the simulation:

```{r estimate_rates}
# Estimate transcription rates
srates <- estimate_simulation_transcription_rates(simpol)

# Get the rates
chi <- get_chi(srates)
beta_org <- get_beta_org(srates)
beta_adp <- get_beta_adp(srates)
phi <- get_phi(srates)
```

## Visualizing Results

The package provides several plotting functions:

```{r plotting}
# Plot transcription rates
plot_transcription_rates(srates)

# Plot pause site distribution
plot_pause_site_distribution(srates)

# Plot RNAP counts (if available)
plot_rnap_counts(srates)
```

# Advanced Usage

## Steric Hindrance

You can enable steric hindrance calculations:

```{r steric_hindrance}
# Estimate rates with steric hindrance
srates_sh <- estimate_simulation_transcription_rates(simpol, steric_hindrance = TRUE)

# Get steric hindrance values
phi_sh <- get_phi(srates_sh)
```

## Saving and Loading Results

You can save your results to RDS files:

```{r save_load}
# Save results
saveRDS(srates, "simulation_results.RDS")

# Load results
srates_loaded <- readRDS("simulation_results.RDS")
```

## Accessing Simulation Parameters

You can access the original simulation parameters:

```{r parameters}
# Get parameters
params <- get_parameters(simpol)

# Print parameters
print(params)
```

# Session Info

```{r session_info}
sessionInfo()
``` 